---
title: "Data Visualization (1)"
author: "Haohan Chen"
date: "Last update: `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: default
  pdf_document: default
  md_document: default
knit: (function(inputFile, encoding){rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = getwd())})
editor_options: 
  chunk_output_type: console
---

## Overview

In this lecture, we start our adventure in Data Visualization with R. Data Visualization is arguably (one of) R's most powerful and irreplaceable functionality.

You can find the concepts and R functionality for data visualization in the following two books.

-   [Kabacoff] Kabacoff, Rob. *Data Visualization with R*. 2020. E-book: [rkabacoff.github.io/datavis](rkabacoff.github.io/datavis)
-   [Healy] Healy, Kieran. *Data visualization: a practical introduction*. Princeton University Press, 2018. E-book: [socviz.co](socviz.co)

In this lecture, we will cover some basics of data visualization.

## In-Class Exercise

## `ggplot`

## 

```{r}
library(tidyverse)
```

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

```{r}



# Binary

e_civil_war # civil war

# categorical

# V-Dem democracy index (categorized, 3 ways of cat): e_v2x_libdem


# What is the lexical index of democracy in the country?
d |> select(e_lexical_index)

# e_fh_status # FH. Free, partically free, not free.
d |> select(e_fh_status)

# Region indicators: e_regiongeo; e_regionpol; e_regionpol_6C  

# Political competition: e_polcomp 
d |> select(e_polcomp)



```
