---
title: "Sentiment Analysis"
author: "Haohan Chen (HKU)"
date: "`r Sys.Date()`"
output: html_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This notebook demonstrate Sentiment Analsyis

```{r}
library(tidyverse)
library(lubridate)
```

```{r}
d_fulltext = read_rds("Lec_10/1_Text_Mining/data/data_ce_speech_article.rds")
# Change the date variable to "date" format
d_fulltext = d_fulltext |> mutate(date_of_speech = dmy(date_of_speech))
```

## Tokenization

```{r}
if (!require("tidytext")) install.packages("tidytext")

library(tidytext) # Full introduction: http://tidytextmining.com/
```

```{r}
d_tokenized = d_fulltext |>
  select(uid, date_of_speech, text) |>
  unnest_tokens(word, text)

head(d_tokenized, 20)

```

## Wrangling: Remove Stop Words

```{r}
# Load Stopwords
data("stop_words")

head(stop_words, 20)
```

```{r}
# Remove stopwords
d_tokenized_s = d_tokenized |>
  anti_join(stop_words, by = "word")
# anti_join: whatever appearing in the stop_words dataframe, we remove it.
```

## Wrangling [Optional]: Stemming

```{r}
if (!require(SnowballC)) install.packages("SnowballC")
library(SnowballC)
```

```{r}
d_tokenized_s = d_tokenized_s |>
  mutate(stem = wordStem(word))

head(d_tokenized_s, 20)
```

## Sentiment Analysis

```{r}
if (!require(textdata)) install.packages("textdata")
library(textdata)
```

### Load Sentiment Dictionary

```{r}
dict_afinn = get_sentiments("afinn")
dict_bing = get_sentiments("bing")
dict_nrc = get_sentiments("nrc") 

table(dict_afinn$value)
table(dict_bing$sentiment)
table(dict_nrc$sentiment)

# Learn more https://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm
```

Note, if you run this function for the first time, you will get a prompt in the console asking you to confirm that you are willing to download the sentiment dataset. The prompt looks as follows. Type "1" to install the dictionaries.

```         
Do you want to download:

Name: AFINN-111

URL: http://www2.imm.dtu.dk/pubdb/views/publication_details.php?id=6010

License: Open Database License (ODbL) v1.0

Size: 78 KB (cleaned 59 KB)

Download mechanism: https

1: Yes

2: No
```

### Calculate the Simplest Sentiment Scores

```{r}
# Merge your tokenized documents with the sentiment dictionary
d_tokenized_s_afinn = d_tokenized_s |>
  select(uid, date_of_speech, word) |>
  inner_join(dict_afinn, by = "word")

# Aggregate the sentiment score for each document
d_tokenized_s_afinn_agg = d_tokenized_s_afinn |>
  group_by(uid, date_of_speech) |>
  summarise(sentiment_score = sum(value))

d_tokenized_s_afinn_agg = d_fulltext |>
  select(uid) |>
  left_join(d_tokenized_s_afinn_agg) |>
  mutate(sentiment_score = replace_na(sentiment_score, 0))

# Change of sentiment over time?
d_tokenized_s_afinn_agg |>
  ggplot(aes(x = date_of_speech, y = sentiment_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  labs(
    title = "Sentiment Scores of Hong Kong CE's Speeches and Articles"
  ) +
  xlab("Date") + ylab("Sentiment Scores")


```

```{r}
# To do it better, we can normalize the sentiment scores by document lengths

# Merge your tokenized documents with the sentiment dictionary
d_tokenized_s_afinn = d_tokenized_s |>
  group_by(uid) |> mutate(doc_length = n()) |>
  ungroup() |>
  select(uid, date_of_speech, word, doc_length) |>
  inner_join(dict_afinn, by = "word") |>
  ungroup()

# Aggregate the sentiment score for each document
d_tokenized_s_afinn_agg = d_tokenized_s_afinn |>
  group_by(uid, date_of_speech) |>
  summarise(sentiment_score = sum(value) / mean(doc_length))

d_tokenized_s_afinn_agg = d_fulltext |>
  select(uid) |>
  left_join(d_tokenized_s_afinn_agg) |>
  mutate(sentiment_score = replace_na(sentiment_score, 0))

# Change of sentiment over time?
d_tokenized_s_afinn_agg |>
  ggplot(aes(x = date_of_speech, y = sentiment_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  labs(
    title = "Sentiment Scores of Hong Kong CE's Speeches and Articles"
  ) +
  xlab("Date") + ylab("Sentiment Scores (Normalized)")

```

## Calculate Scores of Emotions

```{r}
dict_nrc

d_tokenized_s_nrc = d_tokenized_s |>
  inner_join(dict_nrc, by = "word", multiple = "all")

d_tokenized_s_nrc_agg = d_tokenized_s_nrc |>
  group_by(uid, date_of_speech, sentiment) |>
  count() |>
  pivot_wider(names_from = "sentiment", values_from = "n", 
              names_prefix = "sentiment_score_")

names(d_tokenized_s_nrc_agg)

# Change of sentiment over time?
d_tokenized_s_nrc_agg |>
  ggplot(aes(x = date_of_speech, y = sentiment_score_sadness)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  labs(
    title = "Sentiment Scores of Hong Kong CE's Speeches and Articles"
  ) +
  xlab("Date") + ylab("Sadness Scores")
```

```{r}
# Normalize the sentiment scores
d_tokenized_s_nrc = d_tokenized_s |>
  group_by(uid) |>
  mutate(doc_length = n()) |>
  ungroup() |>
  inner_join(dict_nrc, by = "word", multiple = "all")

d_tokenized_s_nrc_agg = d_tokenized_s_nrc |>
  group_by(uid, date_of_speech, sentiment) |>
  summarise(n = n() / mean(doc_length)) |>
  pivot_wider(names_from = "sentiment", values_from = "n", 
              names_prefix = "sentiment_score_")


# Change of sentiment over time?
d_tokenized_s_nrc_agg |>
  ggplot(aes(x = date_of_speech, y = sentiment_score_sadness)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  labs(
    title = "Sentiment Scores of Hong Kong CE's Speeches and Articles"
  ) +
  xlab("Date") + ylab("Sadness Scores (Normalized)")

```
