---
title: "Spatial Data Analysis (II)"
author: "Haohan Chen (HKU)"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.height=10, fig.width=15)
```

## Introduction

This note project folder demonstrate the basics of spatial data analysis with R.

## Required Package

```{r}
library(tidyverse)
library(ggthemes)
theme_set(theme_map())
```

## Load the Conflict Data

```{r}
d = read_csv("Lec_09/data/GEDEvent_v22_1.csv")
names(d)
```

## Map Inter-Country Variation

Draw a world map featuring the variation of conflicts among countries in the year of 2021.

### Put together your data

Create a dataframe named `d_country_2021` that summarizes the number of conflicts in each country in the year of 2021.

```{r}
d_country_2021 = d |>
  filter(year == 2021) |>
  group_by(country, country_id, region) |>
  summarise(
    n_conflict = n()
  )

d_country_2021 = d_country_2021 |> arrange(-n_conflict)

```

## Load the World Map

```{r}
world = map_data("world")

names(world)

# Summary statistics
summary(world)

# Problem: Max. longitude > 180! Remove those
world = world |> filter(long <= 180)
```

```{r}
# Plot the world map
world |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region),
    color = "white", fill = "gray", size = 0.1
  ) +
  coord_map()

```

## Merge Data with the World Map

```{r}
world_m = world |>
  left_join(d_country_2021, by = c("region" = "country"))

```

Check what the merged data look like.

## Now, Visualize Conflict Data on a Map

Color the countries with the number of conflicts in 2021.

```{r}
world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "black", size = 0.1
  ) +
  coord_map()
```

## Make the Map More Informative

Our next step is making the above map more informative

### Change the color scheme

Try the `scale_fill_viridis_c()` function. This should give you the most reliable color scheme.

+-----------+------------------------------------------------------------------------------------------------------------------------------------------------+
| `option`  | A character string indicating the color map option to use. Eight options are available:                                                        |
|           |                                                                                                                                                |
|           | -   "magma" (or "A")                                                                                                                           |
|           |                                                                                                                                                |
|           | -   "inferno" (or "B")                                                                                                                         |
|           |                                                                                                                                                |
|           | -   "plasma" (or "C")                                                                                                                          |
|           |                                                                                                                                                |
|           | -   "viridis" (or "D")                                                                                                                         |
|           |                                                                                                                                                |
|           | -   "cividis" (or "E")                                                                                                                         |
|           |                                                                                                                                                |
|           | -   "rocket" (or "F")                                                                                                                          |
|           |                                                                                                                                                |
|           | -   "mako" (or "G")                                                                                                                            |
|           |                                                                                                                                                |
|           | -   "turbo" (or "H")                                                                                                                           |
+-----------+------------------------------------------------------------------------------------------------------------------------------------------------+
| direction | Sets the order of colors in the scale. If 1, the default, colors are ordered from darkest to lightest. If -1, the order of colors is reversed. |
+-----------+------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}
world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "A") +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B") +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map()

world_m |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = n_conflict),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "D", direction = -1) +
  coord_map()
```

Try the other specifications of color schemes.

### Change the legend's name

The most robust way to change the legend's name is the rename the variable.

Note: When your variable name contains space, you should put "\`" around it.

```{r}
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = `Number of Conflicts`),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map()
```

### Transforming the Scale or Data

The distribution of the conflict frequencies is so skewed that makes the differences of low-conflict countries indistinguishable. Transform the data or the scales to take into account the skewed data.

```{r}
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = `Number of Conflicts`),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1, trans = "log") +
  coord_map()
```

Or we can transform the data

```{r}
world_m |>
  rename("Number of Conflicts" = "n_conflict") |>
  ggplot(aes(x = long, y = lat)) +
  geom_map(
    map = world,
    aes(map_id = region, fill = log(`Number of Conflicts`+1)),
    color = "white", size = 0.1
  ) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  coord_map() +
  theme(legend.position = "bottom")
```

## Choropleth Maps

The maps we have drawn in this notebook are named "choropleth maps."

"A choropleth map displays divided geographical areas or regions that are colored in relation to a numeric variable."
