---
title: "Spatial Data Analysis (IV)"
subtitle: "Mapping China"
author: "Haohan Chen (HKU)"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

## Required Packages

```{r}
library(tidyverse)
library(ggthemes)
theme_set(theme_map())

library(sf)
sf_use_s2(FALSE)

# Display Chinese text with ggplot
library(showtext)
showtext_auto()
```

## Load China Map

```{r}
# https://github.com/xmc811/mapchina
# install.packages("mapchina")
library(mapchina)
chinamap = china
```

## Plot China Map

```{r}
ggplot(data = chinamap) +
  geom_sf()
```

```{r}
ggplot(data = chinamap) +
  geom_sf(aes(fill = Name_Province)) +
  theme(legend.position = "none")
```

## Subsample

```{r}
hkmap = chinamap |> filter(Name_Province == "香港特别行政区")

ggplot(data = hkmap) +
  geom_sf() +
  geom_sf_label(aes(label = Name_County))

ggplot(data = hkmap) +
  geom_sf()
```

## Merge Geographic Units

```{r}
# Merge into provincial-level units
chinamap_province = chinamap |>
  group_by(Code_Province, Name_Province) |>
  summarise(
    geometry = st_union(geometry)
  ) 

chinamap_province |>
  ggplot() + geom_sf()

chinamap_province |>
  ggplot() + geom_sf() +
  geom_sf_text(aes(label = Name_Province))
```

```{r}
chinamap_prefecture = chinamap |>
  mutate(
    Code_Prefecture = Code_Perfecture,
    Name_Prefecture_t = 
      ifelse(is.na(Name_Perfecture), Name_Province, Name_Perfecture)
  ) |>
  group_by(Code_Perfecture, Name_Prefecture_t) |>
  summarise(
    geometry = st_union(geometry)
    ) 

chinamap_prefecture |>
  ggplot() + geom_sf()
```

## Let's Have Some Fun

Visualize China's Socioeconomic data with maps.

```{r}
d_china_province = readxl::read_excel("Lec_09/data/China_province_data.xlsx")
d_china_prefecture = readxl::read_excel("Lec_09/data/China_prefecture_data.xlsx")

```

### Exercise

-   Merge the provincial-level socioeconomic data with the provincial-level map.
-   Draw Choropleths of socioeconomic indicators of interest.
-   Draw Cartograms of socioeconomic indicators of interest.

Does it change your understanding of the political economy of China?
