---
title: "Spatial Data Analysis (V)"
subtitle: "Network"
author: "Haohan Chen (HKU)"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.height=10, fig.width=15, warning=FALSE)
```

## Introduction

This note project folder demonstrate the basics of spatial data analysis with R.

## Required Package

```{r}
library(tidyverse)
library(ggthemes)
theme_set(theme_map())

library(sf)
sf_use_s2(FALSE)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Load the World Map

Introducing a new format of geo-referenced data.

```{r}
world = ne_countries(scale = "medium", type = "map_units", returnclass = "sf")
```

```{r}
world |> ggplot() + geom_sf()
```

```{r}
names(world)
```

## Download Conflict Dataset

```{r, eval=FALSE}
download.file(
  "https://ucdp.uu.se/downloads/ucdpprio/ucdp-prio-acd-221-csv.zip",
  destfile = "data/ucdp-prio-acd-221-csv.zip")

unzip("Lec_09/data/ucdp-prio-acd-221-csv.zip", exdir = "Lec_09/data")

# For more information: https://ucdp.uu.se/downloads/index.html#armedconflict
```

## Load Conflict Dataset

```{r}
d = read_csv("Lec_09/data/ucdp-prio-acd-221.csv")
```

## Extract Entities of Interest

Extract country names and find the ISO3c codes using the package `countrycode`. This is a "lazy" approach. The most robust approach will be a locally customized dataset of country names.

```{r}
library(countrycode)

# Take a look at the data
# Variables of interest: side_a, side_b

d |>
  select(year, side_a, side_b)

d_t = d |>
  mutate(
    side_a_country = countrycode(side_a, "country.name", "country.name"),
    side_a_iso3c = countrycode(side_a, "country.name", "iso3c"),
    .after = "side_a"
  ) |>
  mutate(
    side_b_country = countrycode(side_b, "country.name", "country.name"),
    side_b_iso3c = countrycode(side_b, "country.name", "iso3c"),
    .after = "side_b"
  )

# Need to check manually if the conversion is successful.
d_t |>
  select(year, side_a, side_a_country, side_a_iso3c, side_b, side_b_country, side_b_iso3c)

# It looks already. Then filter only the country-country conflicts
d_t_s = d_t |>
  filter(!is.na(side_a_country) & !is.na(side_b_country))

# OK. I don't think this dataset is entirely clean yet. 
# If this is a real research project, you need to do more manual check.
# But for demonstration purpose, let's use it as it is.

```

## Get the centroids of each country

To draw the network, we need points representing countries' locations). Using the map we already halv, we find the geographic "centroids" for each country using functions from the `sf` package.

```{r}
# Calculate the countries' "centroids"
country_centroid = st_centroid(world$geometry) 

# Convert the centroids to a dataframe
country_centroid_df = country_centroid |> 
  st_coordinates() |>
  as_tibble() |>
  rename("centroid_xcoord" = "X", "centroid_ycoord" = "Y")

world_t = world |>
  bind_cols(country_centroid_df)

```

To check whether the geogrphic centroids are correctly specified we plot a world map along with the points of centroids. Check if the maps make sense and what strike you as interesting/ unintuitive.

```{r}
world_t |>
  ggplot() +
  geom_sf() +
  geom_point(aes(x = centroid_xcoord, y = centroid_ycoord), size = 0.1)
```

## Merge the world map with the network data

Now we have the locations of each country (using their geographic centroids). Merge them with the conflict dataset

```{r}
# Get a dataset with only countries' identifiers and locations of their centroids
world_t_centroid = world_t |>
  as_tibble() |>
  select(iso_a3, centroid_xcoord, centroid_ycoord) |>
  filter(!is.na(iso_a3)) # There are few entries with no country identifiers. We do not need them.


d_t_s_2 = d_t_s |>
  left_join(world_t_centroid, by = c("side_a_iso3c" = "iso_a3")) |>
  rename("centroid_xcoord_a" = "centroid_xcoord", "centroid_ycoord_a" = "centroid_ycoord") |>
  left_join(world_t_centroid, by = c("side_b_iso3c"= "iso_a3")) |>
  rename("centroid_xcoord_b" = "centroid_xcoord", "centroid_ycoord_b" = "centroid_ycoord")


names(d_t_s_2)

# d_t_s_2 |> select(side_a_iso3c, centroid_xcoord_a, centroid_ycoord_a, side_b_iso3c, centroid_xcoord_b, centroid_ycoord_b) |> View()

```

## Finally, we can plot the map

```{r}
# ggplot() +
#   geom_sf(data = world_t) +
#   geom_point(data = d_t_s_2, aes(x = centroid_xcoord_a, y = centroid_ycoord_a)) +
#   geom_point(data = d_t_s_2, aes(x = centroid_xcoord_b, y = centroid_ycoord_b)) +
#   geom_curve(data = d_t_s_2, aes(x = centroid_xcoord_a, y = centroid_ycoord_a,
#                                  xend = centroid_xcoord_a, yend = centroid_ycoord_b))
# ERROR above!

# Remove countries to self
d_t_s_3 = d_t_s_2 |>
  filter(!(side_a_iso3c == side_b_iso3c))

ggplot() +
  geom_sf(data = world_t) +
  geom_point(data = d_t_s_3, aes(x = centroid_xcoord_a, y = centroid_ycoord_a)) +
  geom_point(data = d_t_s_3, aes(x = centroid_xcoord_b, y = centroid_ycoord_b)) +
  geom_curve(data = d_t_s_3, aes(x = centroid_xcoord_a, 
                                 y = centroid_ycoord_a,
                                 xend = centroid_xcoord_b, 
                                 yend = centroid_ycoord_b), 
             curvature = 0.2, color = "red", alpah = 0.5)

ggplot() +
  geom_sf(data = world_t) +
  geom_text(data = d_t_s_3, aes(x = centroid_xcoord_a, y = centroid_ycoord_a, label = side_a_iso3c)) +
  geom_text(data = d_t_s_3, aes(x = centroid_xcoord_b, y = centroid_ycoord_b, label = side_b_iso3c)) +
  geom_curve(data = d_t_s_3, aes(x = centroid_xcoord_a, 
                                 y = centroid_ycoord_a,
                                 xend = centroid_xcoord_b, 
                                 yend = centroid_ycoord_b), curvature = 0.2, color = "red", alpah = 0.5)

```

## Advanced figure, considering the number of conflicts

```{r}
d_t_s_4 = d_t_s_3 |>
  group_by(side_a_iso3c, side_b_iso3c, 
           centroid_xcoord_a, centroid_ycoord_a, 
           centroid_xcoord_b, centroid_ycoord_b) |>
  count() |> ungroup()

ggplot() +
  geom_sf(data = world_t) +
  geom_point(data = d_t_s_4, aes(x = centroid_xcoord_a, y = centroid_ycoord_a)) +
  geom_point(data = d_t_s_4, aes(x = centroid_xcoord_b, y = centroid_ycoord_b)) +
  geom_curve(data = d_t_s_4, aes(x = centroid_xcoord_a, 
                                 y = centroid_ycoord_a,
                                 xend = centroid_xcoord_b, 
                                 yend = centroid_ycoord_b,
                                 linewidth = n), 
             curvature = 0.2, color = "red", alpha = 0.5)

d_top10_n_conflict <- d_t_s_4 |>
  slice_max(n, n = 6)

ggplot() +
  geom_sf(data = world_t) +
  geom_curve(data = d_t_s_4, aes(x = centroid_xcoord_a, 
                                 y = centroid_ycoord_a,
                                 xend = centroid_xcoord_b, 
                                 yend = centroid_ycoord_b,
                                 linewidth = n,
                                 color = n), 
             curvature = 0.2, alpha = 0.5) +
  scale_color_viridis_c(option = "B", trans = "sqrt") +
  geom_text(data = d_top10_n_conflict, aes(x = centroid_xcoord_a, y = centroid_ycoord_a, label = side_a_iso3c)) +
  geom_text(data = d_top10_n_conflict, aes(x = centroid_xcoord_b, y = centroid_ycoord_b, label = side_b_iso3c))

# Q: Why does the text look weird?

```
