---
title: "Data Visualization (3)"
subtitle: "Intermediate: Categorical Variables"
author: "Haohan Chen"
institute: "POLI3148 Data Science in PPA (The University of Hong Kong)"
date: "Last update: `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  # html_document: default
  beamer_presentation:
    theme: "Goettingen"
    colortheme: "default"
    fonttheme: "default"
  md_document: 
    variant: markdown_github
classoption: "aspectratio=169"
knit: (function(inputFile, encoding){rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = getwd())})
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 3, message = FALSE, warning = FALSE)
```

## Objectives

Master data visualization methods for categorical data with `ggplot`.

-   1 cat.

-   2 cat.

-   1 cat. + 1 quant.

-   1 cat. + 2 quant.

-   2 cat. + 1 quant.

## Reading Materials on Data Visualization

-   [Kabacoff] Kabacoff, Rob. Data Visualization with R. 2020. E-book: [rkabacoff.github.io/datavis](https://rkabacoff.github.io/datavis/)

-   [Healy] Healy, Kieran. Data visualization: a practical introduction. Princeton University Press, 2018. E-book: [socviz.co](https://socviz.co/)

# Housekeeping

## Load Data

\scriptsize

```{r, message=FALSE}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document

d <- readRDS("Lec_08/data/wealth_and_health.rds")
d |> print(n = 3)
```

## Variable Types

Visualization tools to use largely depends on variable types

-   "Quantitative" (Continuous, Count)

    -   GDP per capita

    -   Life expectancy

    -   Population

    -   Infant mortality

-   **Categorical**

    -   **Binary: Binary "democracy" indicator**

    -   **Nominal: Region**

    -   **Ordinal: Lexical Index of Electoral Democracy**

## Recoding Categorical Data

\tiny

```{r}
table(d$region)
```

## Recoding Categorical Data

\tiny

```{r}
d <- d |>
  mutate(
    region = case_match(
      region,
      1 ~ "Western Europe", 2 ~ "Northern Europe", 3 ~ "Southern Europe",
      4 ~ "Eastern Europe", 5 ~ "Northern Africa", 6 ~ "Western Africa",
      7 ~ "Middle Africa", 8 ~ "Eastern Africa", 9 ~ "Southern Africa",
      10 ~ "Western Asia", 11 ~ "Central Asia", 12 ~ "Eastern Asia",
      13 ~ "South-Eastern Asia", 14 ~ "Southern Asia", 15 ~ "Oceania",
      16 ~ "North America", 17 ~ "Central America", 18 ~ "South America", 19 ~ "Caribbean",
      .default = NA))
table(d$region)
```

# Cat. X 1

## Visualize One Categorical Variable

-   Summary Statistics

    -   Count

    -   Proportion

-   Visualization

    -   Bar chart

    -   Needle plot

## Summary Statistics

\scriptsize

```{r}
summary(d$region) # Quick summary
length(d$region) # Number of observations
is.na(d$region) |> sum() # Number of missing values
```

## Summary Statistics (con'd)

\tiny

```{r}
table(d$region, useNA = "always")
# table(d$region, useNA = "always") |> as.data.frame() |> as_tibble()
# d |> group_by(region) |> count() # Equivalent tidyverse solution
```

## Summary Statistics (con'd)

\tiny

```{r}
table(d$region, useNA = "always") |> prop.table()
# d |> group_by(region) |> count() |> ungroup() |> mutate(prop = n / sum(n))
```

## Bar Chart: Default

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = region)) + geom_bar() +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Bar Chart: Re-orient text labels

Can't read the text on the x axis? Re-orient it.

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = region)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) + # Try: angle = 45
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Bar Chart: Flip the vertical and horizontal axes

\tiny

```{r, fig.height=4}
d |> ggplot(aes(y = region)) + geom_bar() +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")
```

## Bar Chart: Order by Frequency (1)

Order from the most to least frequent category

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = fct_infreq(region))) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) + # Try: angle = 45
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Bar Chart: Order by Frequency (2)

Order from the least to most frequent category

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = fct_rev(fct_infreq(region)))) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Bar Chart: Use Your Defined Order

\tiny

```{r}
region_levels <- c(
  "Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe",
  "Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa",
  "Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia",
  "Oceania", 
  "North America", "Central America", "South America", "Caribbean")

d |>
  mutate(region = factor(region, levels = rev(region_levels))) |>
  ggplot(aes(y = region)) + geom_bar() +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")
```

## Calculate Frequencies before Visualizing

In all the previous examples, we let `ggplot` calculate the frequency for us. To allow for more customization, a better way is to calculate the frequencies manually before using `geom_bar`.

\tiny

```{r}
d |> group_by(region) |> summarise(n_obs = n())
```

## Calculate Frequencies before Visualizing

\tiny

```{r, fig.height=4}
d |> group_by(region) |> summarise(n_obs = n()) |>
  ggplot(aes(x = reorder(region, -n_obs), y = n_obs)) + geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")  
```

## Calculate the Frequencies before Visualizing

\tiny

```{r, fig.height=4}
d |> group_by(region) |> summarise(n_obs = n()) |>
  ggplot(aes(y = reorder(region, n_obs), x = n_obs)) + geom_bar(stat = "identity") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")  
```

## Too many categories? Recode and Redo the Visualization

An essential technique for the visualization and analysis of categorical data.

\tiny

```{r}
d <- d |>
  mutate(region_higher = case_match(
    region,
    c("Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe") ~ "Europe",
    c("Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa") ~ "Africa",
    c("Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia") ~ "Asia",
    c("Central America", "South America", "Caribbean") ~ "Latin America",
    .default = region), .before = region)

table(d$region_higher)
```

## Re-do the Visualization

\tiny

```{r, fig.height=4}
d |> group_by(region_higher) |> summarise(n_obs = n()) |>
  ggplot(aes(y = reorder(region_higher, n_obs), x = n_obs)) + geom_bar(stat = "identity") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")
```

# Cat. X 2

## A Second Variable: Democracy

\tiny

```{r, fig.height=4}
table(d$region_higher, useNA = "always")
table(d$democracy_binary, useNA = "always")
```

## Region and Democracy: Stacked Bar Chart

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

THIS IS WRONG. What's wrong? `democracy_binary`, a binary variable, is treated as a quantitative variable.

## Region and Democracy: Stacked Bar Chart (rev. 1)

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = factor(democracy_binary))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")  
```

## Region and Democracy: Stacked Bar Chart (rev. 2)

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")  
```

## Region and Democracy: Visualize Proportion

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |> # Calculate proportion
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Percentage (%)", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

## Annotate the Bar Chart

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n_obs), position = position_stack(vjust = 0.5), size = 2) +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

## Annotate the Bar Chart (Proportion)

\tiny

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |>
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n_obs), position = position_stack(vjust = 0.5), size = 2) +
  labs(y = "Region", x = "Proportion", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

**Note:** I label the *counts*, not the *proportions*, so that readers can get the proportion from the bar charts and the actual number form the labels. How do you like it?

# 1 Cat. + 1 Quant.

## Boxplot

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = region_higher, y = life_expectancy)) + 
  geom_boxplot() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Boxplot Re-oriented

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = region_higher, y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Boxplot Re-ordered by Median

\tiny

```{r, fig.height=4}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Boxplot with Actual Data Points

\tiny

```{r, fig.height=4}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA. Essential for the reordering to work!
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + geom_jitter(position = position_jitter(0.3), alpha = 0.1) + # geom_jitter() does the trick.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Violin Plot

\tiny

```{r, fig.height=4}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() +
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Boxplot + Violin Plot

\tiny

```{r, fig.height=4}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() + geom_boxplot(width = 0.15, fill = "gray") + # Note: Place geom_boxplot after geom_violin.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Colored Histogram

Note: The bars are *stacked*.

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = region_higher)) +
  geom_histogram(bins = 50) + 
  labs(title = "Life Expectancy by Region", fill = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Colored Frequency Ploygon

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, color = region_higher)) +
  geom_freqpoly(bins = 50) + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Colored Density

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = region_higher)) +
  geom_density(alpha = 0.2) + 
  labs(title = "Life Expectancy by Region", fill = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Colored Density (alternative)

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = NULL, color = region_higher)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Use Facets: Split into Columns

Does your figure look too "busy"? Separate them into different sub-figures!

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = NULL)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom") +
  facet_grid(cols = vars(region_higher))
```

## Use Facets: Split into Rows

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = NULL)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom") +
  facet_grid(rows = vars(region_higher))
```

## Use Facets: Flexible Organization

\tiny

```{r, fig.height=4}
d |>
  ggplot(aes(x = life_expectancy, fill = NULL)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom") +
  facet_wrap(~region_higher, nrow = 2)
```

# 1 Cat. + 2 Quant.

## The 2 Quant. Plot We Did Last Time

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot: Categorical Variable as Color

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(color = region_higher), alpha = 0.3, shape = 3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot: Categorical Variable as Shape

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot: Categorical Variable as Color and Shape

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  # scale_color_viridis_d(option = "A") +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot with One Trend Line

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot with Separate Trend Lines

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.1, size = 0.5, stroke = 1) +
  geom_smooth(aes(color = region_higher)) +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

## Use Facets 1

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)") +
  facet_wrap(~region_higher, nrow = 2)
```

## Use Facets 2

\tiny

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)") +
  facet_wrap(~region_higher, nrow = 2, scales = "free")
```

## If you really want a colorful faceted plot...

\tiny

You don't really *need* the different colors though...

```{r, fig.height=4}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)") +
  facet_wrap(~region_higher, nrow = 2, scales = "free") +
  theme(legend.position = "none")
```

# 2 Cat. + 1 Quant.

## Heatmap

\tiny

```{r, fig.width=9, fig.height=4}
d |> filter(year >= 1984) |>
  group_by(region_higher, year) |> summarise(life_expectancy_avg = mean(life_expectancy, na.rm = TRUE)) |> mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = region_higher, fill = life_expectancy_avg)) + geom_tile() +
  theme(axis.text.x = element_text(angle = 45), legend.position = "bottom") +
  labs(x = "Year", y = "Region", fill = "Life Expectancy", title = "Life Expectancy by Region-Year")
```

## Heatmap: Changed Color Palette

\tiny

```{r, fig.width=9, fig.height=4}
d |> filter(year >= 1984) |>
  group_by(region_higher, year) |> summarise(life_expectancy_avg = mean(life_expectancy, na.rm = TRUE)) |> mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = region_higher, fill = life_expectancy_avg)) + geom_tile() +
  theme(axis.text.x = element_text(angle = 45), legend.position = "bottom") +
  scale_fill_viridis_c(option = "A", direction = 1) +
  labs(x = "Year", y = "Region", fill = "Life Expectancy", title = "Life Expectancy by Region-Year")
```

## Facets: Density Plot Matrix `facet_grid`

\tiny

```{r, fig.width=16, fig.height=8}
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year)) +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

## Facets: Density Plot Matrix (Free Scales on x)

\tiny

```{r, fig.width=16, fig.height=8}
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year), scales = "free_x") +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

## Facets: Density Plot Matrix (Free Scales on y)

\tiny

```{r, fig.width=16, fig.height=8}
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year), scales = "free_y") +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

## Facets: Density Plot Matrix (Free Scales on both dimensions)

\tiny

```{r, fig.width=16, fig.height=8}
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year), scales = "free") +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

# Many Quant. & Cat.

## Correlation Matrix

\tiny

```{r, fig.width = 10, fig.height=4}
library(GGally)

d |> filter(year == 2019) |>select(gdppc, life_expectancy, population, region_higher) |>
  mutate(population = log(population)) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "log(Population)", "Region"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")),
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), 
    lower = list(continuous = wrap("smooth_loess", alpha = 0.3, color = "gray")))
```
